Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-24 paper/paper.tex
Knapsack

\subsection{Knapsack}\label{sec:knap}
We show now the formalisation and extraction of \emph{dynamorphisms} for
dynamic programming. We use the
knapsack example from~\cite{HinzeWG15}. Dynamorphisms build a memoisation table
that stores intermediate results, alongside the current computation. The
algebra used to define a dynamorphism can access this memoisation table to
speed up computation.  First, we define memoisation tables in terms of
containers.
\begin{coqcode}
Definition MemoShape : Type := A * Sg.
Definition MemoPos := Pg.
Instance Memo : Cont MemoShape MemoPos := { valid := valid \o pair (snd \o fst) snd }.
Definition Table := LFix Memo.
\end{coqcode}
Memoisation tables are the least fixed point of the container defined by shapes
\coq{A * Sg} and positions \coq{Pg}, given a a container \coq{G : Cont Sg Pg}.
We define a function to insert elements into the memo tables:
\begin{coqcode}
Definition Cons : A * App G Table ~> App Memo Table := (* *)
\end{coqcode}
And two functions, to inspect the head of a table, and remove an
element:
\begin{coqcode}
Definition headT : Table ~> A := (* *)
Definition tailT : Table ~> App G Table := (* *)
\end{coqcode}
These tables map ``paths'' in the least fixed point of \coq{Memo} to elements
of type \coq{A}. For example, if \coq{G} is a list-generating functor, these
paths will be natural numbers. 
Using these definitions, a dynamorphism is defined as follows:
\begin{coqcode}
Definition dyna (a : App G Table ~> A) (c : RCoalg G B) : B ~> A
  := headT \o hylo (l_in \o Cons \o pair a id) c.
\end{coqcode}
Note how, instead of an algebra \coq{App G A ~> A}, the algebra takes a
memo table. The definition of the algebra can use this table to lookup
elements, instead of triggering a further recursive call. Elements are inserted
into the memoisation table by the use of \coq{Cons} to the result of applying
the algebra. The algebra for the dynamorphism looks up the previously computed
elements to produce the result, thus saving the corresponding recursive calls:
\begin{coqcode}
Fixpoint memo_knap table wvs :=
  match wvs with | nil => nil | h :: t =>
      match lookupT (Nat.pred (fst h)) table with
      | Some u => (u + snd h)%sint63 :: memo_knapsack table t
      | None => memo_knapsack table t
      end
  end.
Definition knapsack_alg (wvs : list (nat * int))
  (x : App NatF (Table NatF int)) : int :=
  match x with | MkCont sx kx => match sx with
  | inl tt => fun kx => 0%sint63
  | inr tt => fun kx => let tbl := kx next_elem in max_int 0 (memo_knap tbl wvs)
  end kx end.
Definition knapsackA wvs : App NatF (Table NatF int) ~> int :=  (* ... *)
\end{coqcode}
Given the recursive coalgebra for \coq{nat}, \coq{out_nat}, knapsack is defined
as
\begin{coqcode}
Example knapsack wvs : Ext (dyna (knapsackA wvs) out_nat).
\end{coqcode}
While the extracted code has the expected performance characteristics, Coq's
code extraction mechanism is unable to inline several definitions in this case.
We have manually inlined the extracted code for simplicity. The full extracted
code is available in the artefact accompanying this paper, and an interested
reader can check that simple inlining would produce the following function:
% The reader can check
% in our artefact that the extracted code can be trivially inlined to produce the
% following:
\begin{ocamlcode}
let knapsack wvs x = let (y, _) = (let rec f x0 =
      if x0=0 then
        { lFix_out = {shape = Uint63.of_int (0); cont  = fun _ -> f 0 } }
      else let fn := f (x0-1) in { lFix_out = {
           shape = (max_int (Uint63.of_int (0)) (memo_knap fn wvs), sx);
           cont = fun _ -> fn } }
     in f x).lFix_out.shape in y
\end{ocamlcode}
Note how the recursive calls of \ocaml{f} build the memoisation table, and how
this memoisation table is used to compute the intermediate results in
\ocaml{memo_knap}, which is finally discarded to produce the final result.

>>>>

\subsection{Knapsack}\label{sec:knap}
We show now show the formalisation and extraction of \emph{dynamorphisms} for
dynamic programming using the knapsack example from~\cite{HinzeWG15}. In
general, dynamorphisms build a memoisation table which stores intermediate
results alongside the current computation. The algebra used to define a
dynamorphism can access this memoisation table to speed up computation.
First, we define memoisation tables in terms of containers:
\begin{coqcode}
Instance Memo : Cont (A * Sg) Pg := { valid := valid \o pair (snd \o fst) snd }.
Definition Table := LFix Memo.
\end{coqcode}
Given a type of shapes \coq{Sg} and positions \coq{Pg}, the container
\coq{Memo : Cont (A * Sg) Pg}.  The least fixed-point of which is the least
fixed-point of the

given a container \coq{G : Cont Sg Pg}

We define a function to insert elements into the memo tables:
\begin{coqcode}
Definition Cons : A * App G Table ~> App Memo Table := (* *)
\end{coqcode}
And two functions, to inspect the head of a table, and remove an
element:
\begin{coqcode}
Definition headT : Table ~> A := (* *)
Definition tailT : Table ~> App G Table := (* *)
\end{coqcode}
These tables map ``paths'' in the least fixed point of \coq{Memo} to elements
of type \coq{A}. For example, if \coq{G} is a list-generating functor, these
paths will be natural numbers. 
Using these definitions, a dynamorphism is defined as follows:
\begin{coqcode}
Definition dyna (a : App G Table ~> A) (c : RCoalg G B) : B ~> A
  := headT \o hylo (l_in \o Cons \o pair a id) c.
\end{coqcode}
Note how, instead of an algebra \coq{App G A ~> A}, the algebra takes a
memo table. The definition of the algebra can use this table to lookup
elements, instead of triggering a further recursive call. Elements are inserted
into the memoisation table by the use of \coq{Cons} to the result of applying
the algebra. The algebra for the dynamorphism looks up the previously computed
elements to produce the result, thus saving the corresponding recursive calls:
\begin{coqcode}
Fixpoint memo_knap table wvs :=
  match wvs with | nil => nil | h :: t =>
      match lookupT (Nat.pred (fst h)) table with
      | Some u => (u + snd h)%sint63 :: memo_knapsack table t
      | None => memo_knapsack table t
      end
  end.
Definition knapsack_alg (wvs : list (nat * int))
  (x : App NatF (Table NatF int)) : int :=
  match x with | MkCont sx kx => match sx with
  | inl tt => fun kx => 0%sint63
  | inr tt => fun kx => let tbl := kx next_elem in max_int 0 (memo_knap tbl wvs)
  end kx end.
Definition knapsackA wvs : App NatF (Table NatF int) ~> int :=  (* ... *)
\end{coqcode}
Given the recursive coalgebra for \coq{nat}, \coq{out_nat}, knapsack is defined
as
\begin{coqcode}
Example knapsack wvs : Ext (dyna (knapsackA wvs) out_nat).
\end{coqcode}
While the extracted code has the expected performance characteristics, Coq's
code extraction mechanism is unable to inline several definitions in this case.
We have manually inlined the extracted code for simplicity. The full extracted
code is available in the artefact accompanying this paper, and an interested
reader can check that simple inlining would produce the following function:
% The reader can check
% in our artefact that the extracted code can be trivially inlined to produce the
% following:
\begin{ocamlcode}
let knapsack wvs x = let (y, _) = (let rec f x0 =
      if x0=0 then
        { lFix_out = {shape = Uint63.of_int (0); cont  = fun _ -> f 0 } }
      else let fn := f (x0-1) in { lFix_out = {
           shape = (max_int (Uint63.of_int (0)) (memo_knap fn wvs), sx);
           cont = fun _ -> fn } }
     in f x).lFix_out.shape in y
\end{ocamlcode}
Note how the recursive calls of \ocaml{f} build the memoisation table, and how
this memoisation table is used to compute the intermediate results in
\ocaml{memo_knap}, which is finally discarded to produce the final result.
